---
title: "Diagnostics on meta-analysis models and implications for assessing replicability"
author: "Yefeng Yanga & Shinichi Nakagawa"
date: "02-10-2025"
output:
  rmdformats::downcute:
    code_folding: show
    self_contained: true
    thumbnails: false
    lightbox: true
    downcute_theme: "chaos"
pkgdown:
  as_is: true
---

# Prerequisite

Load all necessary packages. Package information is present at the end of this website.

```{r package, warning=FALSE, echo=TRUE}
set.seed(2024)
suppressPackageStartupMessages({
  library(dplyr)
  library(readr)
  library(tidyr) 
  library(distributional)
  library(here)
  library(readxl)
  library(ggplot2)
  library(metafor)
  library(cowplot)
  library(ggdist)
  library(patchwork)
  library(forcats)
  library(RColorBrewer)
  library(viridis)
  library(ggrepel)}
  )

```

# First example

The first example is a meta-analysis by Møller et al. (2012) comparing extracorporeal versus extracorporeal coronary artery bypass grafting (CABG) with heparin doses of less than 300 mg/k in the reduction of postoperative atrial fibrillation (POAF). We did not collect data directly from Møller et al (2012). Instead, we obtained data from the supplementary material of Xiao et al (2024).

## Data

Load the data provided by Xiao et. al. (2024).

```{r}
# load example data
load(here("data","moller12.rda"))
# make a copy
dat <- moller12
knitr::kable(dfround(dat,3) %>% head(), "pipe")
```

## Fit a RE model

```{r}
# fit
dat <- escalc(measure="OR", ai = r1, n1i = n1, ci = r2, n2i = n2, data = dat)
res <- rma(yi, vi, data = dat, method = 'REML', control = list(stepadj = 0.5, maxiter = 1000))
print(res)
```

## Diagnoise the fitted model

```{r}
# all measures at one go
inf <- influence(res)$inf %>% as.data.frame()
inf$study.name <- as.character(dat$study.name) 
inf$id <- 1:nrow(inf)

knitr::kable(dfround(inf,3) %>% head(), "pipe")
```

## Visualization

Visualize all results.

```{r, warning=FALSE}
# check for NA values and remove them if necessary
inf <- inf %>%
  filter(!is.na(weight) & !is.na(rstudent) & !is.na(dffits) & !is.na(cook.d) & !is.na(study.name))

inf <- inf %>%
  mutate(cov_status = ifelse(cov.r >= 1, "Increased", "Decreased"))

p1 <- inf %>% 
 ggplot(aes(x = dffits, y = rstudent)) +
  geom_hline(yintercept = 1.96, size = 0.5, linetype = "dashed", color = "red") +
  geom_hline(yintercept = -1.96, size = 0.5, linetype = "dashed", color = "red") +
  geom_point(aes(fill = cook.d, size = weight), shape = 21) +
  scale_fill_gradientn(colors = rev(brewer.pal(11, "RdBu"))) +
  scale_y_continuous(limits = c(-3, 5)) +
  labs(caption = "", x = "Changes in model coefficients", y = "Externally standardized residuals", fill = "Cook distance", size = "Weight") +
  theme_bw() +
  theme(
    axis.text = element_text(size = 16, color = "black"), 
    axis.title = element_text(size = 18, color = "black"), 
    legend.position = "top",
    legend.title=element_text(size = 16), 
    legend.text=element_text(size = 16)
  ) + 
   geom_text(aes(label = id), size = 3, hjust = 0.5, vjust = 2)
p1
```


```{r}
p2 <- inf %>% 
 ggplot(aes(x = cov.r, y = rstudent)) +
  geom_hline(yintercept = 1.96, size = 0.5, linetype = "dashed", color = "red") +
  geom_hline(yintercept = -1.96, size = 0.5, linetype = "dashed", color = "red") +
  geom_point(aes(fill = QE.del, size = hat), shape = 21) +
  scale_fill_gradientn(breaks = c(45, 50, 65, 80), colors = viridis(n = 10)) +
  scale_y_continuous(limits = c(-3, 5)) +
  labs(caption = "", x = "Changes in variance", y = "Externally standardized residuals", fill = "Cochran's Q", size = "Hat value") +
  theme_bw() +
  theme(
    axis.text = element_text(size = 16, color = "black"), 
    axis.title = element_text(size = 18, color = "black"), 
    legend.position = "top",
    legend.title=element_text(size = 16), 
    legend.text=element_text(size = 16)
  ) + 
   geom_text(aes(label = id), size = 3, hjust = 0.5, vjust = 3)
p2
```


# Second example

The second example is a meta-anlaysis by Colditz et. al. (1994) examining the effectivess of the bacillus Calmette-Guerin (BCG) vaccine for preventing tuberculosis. It is publicly available at `metafor` package.

## Data

Load data from `metafor` package

```{r}
dat <- dat.bcg
dat <- escalc(measure = "RR", ai = tpos, bi = tneg, ci = cpos, di = cneg, data = dat)
knitr::kable(dfround(dat,3) %>% head(), "pipe")
```

## Fit a RE model

```{r}
res <- rma(yi, vi, mods = ~ ablat + year, method = 'REML', data = dat, control = list(stepadj = 0.5, maxiter = 1000))
print(res)
```

## Diagnoise the fitted model

```{r}
# all measures at one go
inf <- influence(res)$inf %>% as.data.frame()
inf$study.name <- as.character(dat$author) 
inf$id <- 1:nrow(inf)
knitr::kable(dfround(inf,3) %>% head(), "pipe")
```

## Visualization

```{r, warning=FALSE}
p3 <- inf %>% 
 ggplot(aes(x = weight, y = rstudent)) +
  geom_hline(yintercept = 1.96, size = 0.5, linetype = "dashed", color = "red") +
  geom_hline(yintercept = -1.96, size = 0.5, linetype = "dashed", color = "red") +
  geom_point(aes(size = cook.d, fill = dffits), shape = 21) +
  scale_fill_gradientn(breaks = c(-3, -2, -1, 0, 1), labels = c(-3, -2, -1, 0, 1), colors = rev(brewer.pal(11, "RdBu"))) +
  scale_y_continuous(limits = c(-3, 3)) +
  labs(caption = "", x = "Weight", y = "Externally standardized residuals", fill = "Coefficient difference", size = "Cook distance") +
  theme_bw() +
  theme(
    axis.text = element_text(size = 16, color = "black"), 
    axis.title = element_text(size = 18, color = "black"), 
    legend.position = "right",
    legend.title=element_text(size = 16), 
    legend.text=element_text(size = 16)
  ) + 
   geom_text(aes(label = id), size = 6, hjust = 0.5, vjust = 2) 

p3
```

# Standard procedures of model diagonistics

We also use the example of Møller et al. (2012) to illustrate the Standard procedures of model diagonistics of meta-analysis.

`metafor` provides all diagonistic measures in one-go.

```{r}
# load example data
load(here("data","moller12.rda"))
# make a copy
dat <- moller12
# fit
dat <- escalc(measure="OR", ai = r1, n1i = n1, ci = r2, n2i = n2, data = dat)
res <- rma(yi, vi, data = dat, method = 'REML', control = list(stepadj = 0.5, maxiter = 1000)) 
res.l1o <- leave1out(res) %>% as.data.frame()
res.l1o$study.name <- as.character(dat$study.name) 
res.l1o$id <- 1:nrow(res.l1o)
knitr::kable(dfround(res.l1o,3) %>% head(), "pipe")
```

### Beta coefficient

```{r, warning=FALSE}
# visualization
## beta
p.beta.del <- ggplot(res.l1o) +
  geom_hline(yintercept = res$ci.lb, lty = 3, lwd = 0.75, colour = "lightgrey") +
  geom_hline(yintercept = res$b[1], lty = 1, lwd = 0.75, colour = "grey50") +
  geom_hline(yintercept = res$ci.ub, lty = 3, lwd = 0.75, colour = "lightgrey") +
  geom_pointrange(aes(x = id, y = estimate, ymin = ci.lb, ymax = ci.ub), color = "#E6AB02") +
  scale_x_continuous(limits = c(1,17), breaks = seq(1,17,1), labels = seq(1,17,1)) +
  xlab("Study left out") + 
  ylab("Overall mean effect size (95% CI)") + 
  coord_flip() +
  theme(panel.grid.minor = element_blank())+
  theme_bw() + 
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line.x = element_line(colour = "black")) +
  theme(axis.text = element_text(size = 16, color = "black"), 
    axis.title = element_text(size = 18, color = "black"))
p.beta.del
```


### Between-study variance

```{r, warning=FALSE}
## tau2
p.tau2.del <- res.l1o %>% 
  ggplot(aes(x = id, y = tau2)) +
  geom_segment(aes(x = id, xend = id), 
               yend = 0, size = 1.2, alpha = 0.8, color = "grey60") +
  geom_point(shape = 21, aes(fill = tau2, size = tau2)) +
  scale_fill_gradientn(colors = viridis(n = 10)) +
  scale_x_continuous(limits = c(1,17), breaks = seq(1,17,1), labels = seq(1,17,1)) + # expand = c(0, 0)
  scale_y_continuous(limits = c(0, 1)) +
  coord_flip() +
  labs(x = "Study left out", y = expression(paste("Estimated between-study variance ", tau^2)), fill = expression(tau^2)) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text = element_text(size = 16, color = "black"), 
        axis.title = element_text(size = 18, color = "black"), 
        legend.position = "none",
        legend.title = element_text(size = 14), 
        legend.text = element_text(size = 14)
  )
p.tau2.del
```

### Heterogeiety quantification

```{r, warning=FALSE}
## I2
p.i2.del <- res.l1o %>% 
  ggplot(aes(x = id, y = I2)) +
  geom_segment(aes(x = id, xend = id), 
               yend = 0, size = 1.2, alpha = 0.8, color = "grey60") +
  geom_point(shape = 21, aes(fill = I2, size = I2)) +
  scale_fill_gradientn(colors = viridis(n = 10)) +
  scale_x_continuous(limits = c(1,17), breaks = seq(1,17,1), labels = seq(1,17,1)) + # expand = c(0, 0)
  scale_y_continuous(limits = c(0, 100)) +
  coord_flip() +
  labs(x = "Study left out", y = expression(paste("Heterogeneity ", I^2)), fill = expression(I^2)) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text = element_text(size = 16, color = "black"), 
        axis.title = element_text(size = 18, color = "black"), 
        legend.position = "none",
        legend.title = element_text(size = 14), 
        legend.text = element_text(size = 14)
  )

p.i2.del
```

### Heterogeneity test

```{r, warning=FALSE}
## Q
p.Q.del <- res.l1o %>% 
  ggplot(aes(x = id, y = Q)) +
  geom_segment(aes(x = id, xend = id), 
               yend = 0, size = 1.2, alpha = 0.8, color = "grey60") +
  geom_point(shape = 21, aes(fill = Q, size = Q)) +
  scale_fill_gradientn(colors = viridis(n = 10)) +
  scale_x_continuous(limits = c(1,17), breaks = seq(1,17,1), labels = seq(1,17,1)) + # expand = c(0, 0)
  scale_y_continuous(limits = c(0, 100)) +
  coord_flip() +
  labs(x = "Study left out", y = expression(paste("Heterogeneity Cochran’s Q"))) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text = element_text(size = 16, color = "black"), 
        axis.title = element_text(size = 18, color = "black"), 
        legend.position = "none",
        legend.title = element_text(size = 14), 
        legend.text = element_text(size = 14)
  )

p.Q.del
```

# Package information

```{r}
subset(data.frame(sessioninfo::package_info()), attached==TRUE, c(package, loadedversion))
```

