---
title: "Untitled"
output: html_document
date: '2024-08-10'
---

# Prerequisite

```{r package, warning=FALSE, echo=TRUE}
set.seed(2024)
suppressPackageStartupMessages({
  library(dplyr)
  library(readr)
  library(tidyr) 
  library(distributional)
  library(here)
  library(readxl)
  library(ggplot2)
  library(metafor)
  library(cowplot)
  library(ggdist)
  library(patchwork)
  library(forcats)
  library(RColorBrewer)
  library(viridis)
  library(ggrepel)}
  )

```

# First example

## Data

```{r}
# load example data
load(here("data","moller12.rda"))
# make a copy
dat <- moller12
```

## Fit a RE model
```{r}
# fit
dat <- escalc(measure="OR", ai = r1, n1i = n1, ci = r2, n2i = n2, data = dat)
res <- rma(yi, vi, data = dat, method = 'REML', control = list(stepadj = 0.5, maxiter = 1000)) # https://github.com/cran/metafor/blob/master/R/influence.rma.uni.r
# all measures at one go
inf <- influence(res)$inf %>% as.data.frame()
inf$study.name <- as.character(dat$study.name) 
inf$id <- 1:nrow(inf)

```

## Visualization

https://github.com/cxli233/FriendsDontLetFriends#1-friends-dont-let-friends-make-bar-plots-for-means-separation

```{r}
# check for NA values and remove them if necessary
inf <- inf %>%
  filter(!is.na(weight) & !is.na(rstudent) & !is.na(dffits) & !is.na(cook.d) & !is.na(study.name))

inf <- inf %>%
  mutate(cov_status = ifelse(cov.r >= 1, "Increased", "Decreased"))

p1 <- inf %>% 
 ggplot(aes(x = dffits, y = rstudent)) +
  geom_hline(yintercept = 1.96, size = 0.5, linetype = "dashed", color = "red") +
  geom_hline(yintercept = -1.96, size = 0.5, linetype = "dashed", color = "red") +
  geom_point(aes(fill = cook.d, size = weight), shape = 21) +
  scale_fill_gradientn(colors = rev(brewer.pal(11, "RdBu"))) +
  scale_y_continuous(limits = c(-3, 5)) +
  labs(caption = "", x = "Coefficient difference", y = "Externally standardized residuals", fill = "Cook distance", size = "Weight") +
  theme_bw() +
  theme(
    axis.text = element_text(size = 16, color = "black"), 
    axis.title = element_text(size = 18, color = "black"), 
    legend.position = "top",
    legend.title=element_text(size = 16), 
    legend.text=element_text(size = 16)
  ) + 
   geom_text(aes(label = id), size = 3, hjust = 0.5, vjust = 2)  
   #geom_text(aes(x = 6.54233 + 0.6, y = 4.841277, size = 4, label = "MASS III 2009"))


p2 <- inf %>% 
 ggplot(aes(x = cov.r, y = rstudent)) +
  geom_hline(yintercept = 1.96, size = 0.5, linetype = "dashed", color = "red") +
  geom_hline(yintercept = -1.96, size = 0.5, linetype = "dashed", color = "red") +
  geom_point(aes(fill = QE.del, size = hat), shape = 21) +
  scale_fill_gradientn(breaks = c(45, 50, 65, 80), colors = viridis(n = 10)) +
  scale_y_continuous(limits = c(-3, 5)) +
  labs(caption = "", x = "Ratio of variance", y = "Externally standardized residuals", fill = "Cochran's Q", size = "Hat value") +
  theme_bw() +
  theme(
    axis.text = element_text(size = 16, color = "black"), 
    axis.title = element_text(size = 18, color = "black"), 
    legend.position = "top",
    legend.title=element_text(size = 16), 
    legend.text=element_text(size = 16)
  ) + 
   geom_text(aes(label = id), size = 3, hjust = 0.5, vjust = 3)


png(filename = "fig 3.png", width = 10, height = 12, units = "in", type = "windows", res = 800)
p1 + p2 + plot_layout(nrow = 2, ncol = 1) +
  plot_annotation(tag_levels = 'A') & 
  theme(plot.tag = element_text(size = 18, color = "black"))
dev.off() 

```




# Third example

## Data

```{r}
dat3 <- escalc(measure = "RR", ai = tpos, bi = tneg, ci = cpos, di = cneg, data = dat.bcg)
```


## Fit a RE model

```{r}


res3 <- rma(yi, vi, mods = ~ ablat + year, method = 'REML', data = dat3, control = list(stepadj = 0.5, maxiter = 1000))
# all measures at one go
inf3 <- influence(res3)$inf %>% as.data.frame()
inf3$study.name <- as.character(dat3$author) 
inf3$id <- 1:nrow(inf3)
```

## Visualization

```{r}
p3 <- inf3 %>% 
 ggplot(aes(x = weight, y = rstudent)) +
  geom_hline(yintercept = 1.96, size = 0.5, linetype = "dashed", color = "red") +
  geom_hline(yintercept = -1.96, size = 0.5, linetype = "dashed", color = "red") +
  geom_point(aes(size = cook.d, fill = dffits), shape = 21) +
  scale_fill_gradientn(breaks = c(-3, -2, -1, 0, 1), labels = c(-3, -2, -1, 0, 1), colors = rev(brewer.pal(11, "RdBu"))) +
  scale_y_continuous(limits = c(-3, 3)) +
  labs(caption = "", x = "Weight", y = "Externally standardized residuals", fill = "Coefficient difference", size = "Cook distance") +
  theme_bw() +
  theme(
    axis.text = element_text(size = 16, color = "black"), 
    axis.title = element_text(size = 18, color = "black"), 
    legend.position = "right",
    legend.title=element_text(size = 16), 
    legend.text=element_text(size = 16)
  ) + 
   geom_text(aes(label = id), size = 6, hjust = 0.5, vjust = 2)  


```

# Standard practices of model diagonistics

## Leave-one-out


```{r}
# load example data
load(here("data","moller12.rda"))
# make a copy
dat <- moller12
# fit
dat <- escalc(measure="OR", ai = r1, n1i = n1, ci = r2, n2i = n2, data = dat)
res <- rma(yi, vi, data = dat, method = 'REML', control = list(stepadj = 0.5, maxiter = 1000)) 
res.l1o <- leave1out(res) %>% as.data.frame()
res.l1o$study.name <- as.character(dat$study.name) 
res.l1o$id <- 1:nrow(res.l1o)

# visualization
## beta
p.beta.del <- ggplot(res.l1o) +
  #geom_hline(yintercept = 0, lty = 2, lwd = 0.75, colour = "lightgrey") +
  geom_hline(yintercept = res$ci.lb, lty = 3, lwd = 0.75, colour = "lightgrey") +
  geom_hline(yintercept = res$b[1], lty = 1, lwd = 0.75, colour = "grey50") +
  geom_hline(yintercept = res$ci.ub, lty = 3, lwd = 0.75, colour = "lightgrey") +
  geom_pointrange(aes(x = id, y = estimate, ymin = ci.lb, ymax = ci.ub), color = "#E6AB02") +
  scale_x_continuous(limits = c(1,17), breaks = seq(1,17,1), labels = seq(1,17,1)) +
  xlab("Study left out") + 
  ylab("Overall mean effect size (95% CI)") + 
  coord_flip() +
  theme(panel.grid.minor = element_blank())+
  theme_bw() + 
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        #panel.border = element_blank(),
        #axis.ticks.y = element_blank(),
        axis.line.x = element_line(colour = "black")) +
  theme(axis.text = element_text(size = 16, color = "black"), 
    axis.title = element_text(size = 18, color = "black"))

## tau2
p.tau2.del <- res.l1o %>% 
  ggplot(aes(x = id, y = tau2)) +
  geom_segment(aes(x = id, xend = id), 
               yend = 0, size = 1.2, alpha = 0.8, color = "grey60") +
  geom_point(shape = 21, aes(fill = tau2, size = tau2)) +
  scale_fill_gradientn(colors = viridis(n = 10)) +
  scale_x_continuous(limits = c(1,17), breaks = seq(1,17,1), labels = seq(1,17,1)) + # expand = c(0, 0)
  scale_y_continuous(limits = c(0, 1)) +
  coord_flip() +
  labs(x = "Study left out", y = expression(paste("Estimated between-study variance ", tau^2)), fill = expression(tau^2)) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text = element_text(size = 16, color = "black"), 
        axis.title = element_text(size = 18, color = "black"), 
        legend.position = "none",
        legend.title = element_text(size = 14), 
        legend.text = element_text(size = 14)
  )

## I2
p.i2.del <- res.l1o %>% 
  ggplot(aes(x = id, y = I2)) +
  geom_segment(aes(x = id, xend = id), 
               yend = 0, size = 1.2, alpha = 0.8, color = "grey60") +
  geom_point(shape = 21, aes(fill = I2, size = I2)) +
  scale_fill_gradientn(colors = viridis(n = 10)) +
  scale_x_continuous(limits = c(1,17), breaks = seq(1,17,1), labels = seq(1,17,1)) + # expand = c(0, 0)
  scale_y_continuous(limits = c(0, 100)) +
  coord_flip() +
  labs(x = "Study left out", y = expression(paste("Heterogeneity ", I^2)), fill = expression(I^2)) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text = element_text(size = 16, color = "black"), 
        axis.title = element_text(size = 18, color = "black"), 
        legend.position = "none",
        legend.title = element_text(size = 14), 
        legend.text = element_text(size = 14)
  )

## Q
p.Q.del <- res.l1o %>% 
  ggplot(aes(x = id, y = Q)) +
  geom_segment(aes(x = id, xend = id), 
               yend = 0, size = 1.2, alpha = 0.8, color = "grey60") +
  geom_point(shape = 21, aes(fill = Q, size = Q)) +
  scale_fill_gradientn(colors = viridis(n = 10)) +
  scale_x_continuous(limits = c(1,17), breaks = seq(1,17,1), labels = seq(1,17,1)) + # expand = c(0, 0)
  scale_y_continuous(limits = c(0, 100)) +
  coord_flip() +
  labs(x = "Study left out", y = expression(paste("Heterogeneity Cochran’s Q"))) +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text = element_text(size = 16, color = "black"), 
        axis.title = element_text(size = 18, color = "black"), 
        legend.position = "none",
        legend.title = element_text(size = 14), 
        legend.text = element_text(size = 14)
  )


png(filename = "fig 1.png", width = 12, height = 12, units = "in", type = "windows", res = 800)
p.beta.del +  p.tau2.del + p.i2.del + p.Q.del + plot_layout(nrow = 2, ncol = 2) +
  plot_annotation(tag_levels = 'A') & 
  theme(plot.tag = element_text(size = 18, color = "black"))
dev.off() 


```


# Estimator

https://www.metafor-project.org/doku.php/analyses:dersimonian2007

https://www.metafor-project.org/doku.php/analyses:viechtbauer2007a

https://www.metafor-project.org/doku.php/analyses:viechtbauer2005

Alternative meta-analysis of behavioral interventions to promote action on climate change yields different conclusions

A comparison of heterogeneity variance estimators in simulated random‐effects meta‐analyses

```{r}
# load data
van_der_Linden <- read_xlsx(here("data", "van_der_Linden.xlsx"))

# only select the relevant variables
dat <- data.frame(Study = van_der_Linden$Study, yi = van_der_Linden$SMD, vi = (van_der_Linden$se)^2)

# different estimators
res.PM  <- rma(yi, vi, method = "PM", data = dat)
res.CA  <- rma(yi, vi, method = "HE", data = dat)
res.DL  <- rma(yi, vi, method = "DL", data = dat)
res.CA2 <- rma(yi, vi, method = "GENQ", weights = 1/(vi + res.CA$tau2), data = dat)
res.DL2 <- rma(yi, vi, method = "GENQ", weights=1/(vi + res.DL$tau2), data = dat)
res.CA2 <- rma(yi, vi, tau2 = res.CA2$tau2, data = dat)
res.DL2 <- rma(yi, vi, tau2 = res.DL2$tau2, data = dat)
res.EB   <- rma(yi, vi, method = "EB",   data = dat)
res.ML   <- rma(yi, vi, method = "ML",   data = dat)
res.REML <- rma(yi, vi, method = "REML", data = dat)
res.HS   <- rma(yi, vi, method = "HS",   data = dat)
res.SJ   <- rma(yi, vi, method = "SJ",   data = dat)
res.SJ2  <- rma(yi, vi, method = "SJ",   data = dat, control = list(tau2.init = res.CA$tau2))

# dataframe
res.list <- list(res.PM, res.CA, res.DL, res.CA2, res.DL2, res.EB, 
                res.ML, res.REML, res.HS, res.SJ, res.SJ2)

df.est <- data.frame(method = c("PM", "CA", "DL", "CA2", "DL2", "EB", "ML", "REML", "HS", "SJ", "SJ2"),
  mu  = sapply(res.list, coef),
  ci.lb  = sapply(res.list, function(x) x$ci.lb),
  ci.ub  = sapply(res.list, function(x) x$ci.ub),
  se = sapply(res.list, function(x) x$se[1]), 
  tau2 = sapply(res.list, function(x) x$tau2)) %>% dfround(3)
df.est$method <- as.factor(df.est$method)
df.est$method <- factor(df.est$method, levels = c("SJ2", "SJ", "HS", "REML", "ML", "EB", "DL2", "CA2", "DL", "CA", "PM"))

# plot
# https://mjskay.github.io/ggdist/articles/freq-uncertainty-vis.html
p.est <- df.est %>% 
  ggplot(aes(y = method, fill = method)) +
  stat_halfeye(
    aes(xdist = dist_normal(mu = mu, sigma = se)), size = 2, stroke = 1) +
  theme_bw() +
  labs(x = "Overall mean effect size (95% CI)", y = expression(paste("Estimator"))) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
    axis.text = element_text(size = 16, color = "black"), 
    axis.title = element_text(size = 18, color = "black"), 
    legend.position = "none",
    legend.title = element_text(size = 16), 
    legend.text = element_text(size = 16)
  )

png(filename = "fig 2.png", width = 9, height = 9, units = "in", type = "windows", res = 800)
p.est
dev.off() 

```



# knha

```{r}
# adjustment
res2.PM  <- rma(yi, vi, method = "PM", test = "knha", data = dat)
res2.CA  <- rma(yi, vi, method = "HE", test = "knha", data = dat)
res2.DL  <- rma(yi, vi, method = "DL", test = "knha", data = dat)
 
res2.CA2 <- rma(yi, vi, method = "GENQ", weights = 1/(vi + res.CA$tau2), test = "knha", data = dat)
res2.DL2 <- rma(yi, vi, method = "GENQ", weights=1/(vi + res.DL$tau2), test = "knha", data = dat)
res2.CA2 <- rma(yi, vi, tau2 = res.CA2$tau2, test = "knha", data = dat)
res2.DL2 <- rma(yi, vi, tau2 = res.DL2$tau2, test = "knha", data = dat)
 
res2.EB   <- rma(yi, vi, method = "EB", test = "knha", data = dat)
res2.ML   <- rma(yi, vi, method = "ML", test = "knha", data = dat)
res2.REML <- rma(yi, vi, method = "REML", data = dat)
res2.HS   <- rma(yi, vi, method = "HS", test = "knha", data = dat)
res2.SJ   <- rma(yi, vi, method = "SJ", test = "knha", data = dat)
res2.SJ2  <- rma(yi, vi, method = "SJ", test = "knha", data = dat, control = list(tau2.init = res.CA$tau2))


# extraction
res.list2 <- list(res2.PM, res2.CA, res2.DL, res2.CA2, res2.DL2, res2.EB, 
                res2.ML, res2.REML, res2.HS, res2.SJ, res2.SJ2)

df.est2 <- data.frame(method = c("PM", "CA", "DL", "CA2", "DL2", "EB", "ML", "REML", "HS", "SJ", "SJ2"),
  mu  = sapply(res.list2, coef),
  ci.lb  = sapply(res.list2, function(x) x$ci.lb),
  ci.ub  = sapply(res.list2, function(x) x$ci.ub),
  se = sapply(res.list2, function(x) x$se[1]),
  tau2 = sapply(res.list2, function(x) x$tau2)) %>% dfround(3)


```
